---
title: "Introduction to classification (STAT5003 2016)"
author: "Pengyi Yang"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

## Create sumulation dataset
```{r}
# create positive class sample with 2 descriptive features
set.seed(3)
f1 <- rnorm(100, mean=6, sd = 1.2)
set.seed(4)
f2 <- rnorm(100, mean=6, sd = 1.2)
P.data <- cbind(f1, f2)

# create positive class sample with 2 descriptive features
set.seed(7)
f1 <- rnorm(300, mean=4, sd = 1.2)
set.seed(8)
f2 <- rnorm(300, mean=4, sd = 1.2)
N.data <- cbind(f1, f2)

# combine all samples
data.mat <- data.frame(rbind(P.data, N.data), Class=rep(c(1, 0), time=c(nrow(P.data), nrow(N.data))))
rownames(data.mat) <- paste("s", 1:(nrow(P.data)+nrow(N.data)), sep="")

# plot data
plot(subset(data.mat, Class==1)[,-3], col="red", pch=16, ylim=c(0, 9), xlim=c(0, 9), xlab="Feature 1", ylab="Feature 2")
points(subset(data.mat, Class==0)[,-3], col="blue", pch=16)
```

## Create and utilise logistic regression model for data classification
```{r}
# train a logistic regression model
logit.model <- glm(Class~., family=binomial(link='logit'), data=data.mat)
# plot fitted values from logistic regression model
plot(logit.model$fitted.values)
# use fitted value to classify samples
logit.decision <- ifelse(logit.model$fitted.values > 0.5, 1, 0)
# calculate classification accuracy (in percentage %)
sum(logit.decision == data.mat$Class) / nrow(data.mat) * 100
```


## Create and utilise linear discriminant analysis (LDA) model for data classification
```{r}
library(MASS)
# train an LDA model
lda.model <- lda(Class~., data=data.mat)
lda.fitted <- predict(lda.model, data.mat)$posterior[,"1"]
# plot fitted values from LDA model
plot(lda.fitted)
# use fitted value to classify samples
lda.decision <- ifelse(lda.fitted > 0.5, 1, 0)
# calculate classification accuracy (in percentage %)
sum(lda.decision == data.mat$Class) / nrow(data.mat) * 100
```

## compare Logistic regression classification with LDA classification
```{r}
cor(logit.model$fitted.values, lda.fitted)
plot(logit.model$fitted.values, lda.fitted)
```

## Create and utilise kNN model for data classification
```{r}
library(class)
# a knn with k=1
knn.model1 <- knn(train=data.mat[,-3], test=data.mat[,-3], cl=data.mat[,3], k=1)
# a knn with k=5
knn.model2 <- knn(train=data.mat[,-3], test=data.mat[,-3], cl=data.mat[,3], k=5)
# a knn with k=50
knn.model3 <- knn(train=data.mat[,-3], test=data.mat[,-3], cl=data.mat[,3], k=50)

# calculate classification accuracy
sum(knn.model1 == data.mat$Class) / nrow(data.mat) * 100
sum(knn.model2 == data.mat$Class) / nrow(data.mat) * 100
sum(knn.model3 == data.mat$Class) / nrow(data.mat) * 100

# apply knn and enable calculation of prediction probability
knn.model <- knn(train=data.mat[,-3], test=data.mat[,-3], cl=data.mat[,3], k=5, prob=TRUE)
# extract prediction probability from the prediction model
knn.prob <- attr(knn.model, "prob")
isNegativeSample <- data.mat[,3] != 1
knn.prob[isNegativeSample] <-  1 - knn.prob[isNegativeSample]

plot(knn.prob)
```

## compare kNN classification with Logistic regression classification and LDA classification
```{r}
# combine classification results into a data frame
classifications <- data.frame(logit=logit.model$fitted.values, lda=lda.fitted, knn=knn.prob)
# calculate correlation
cor(classifications)
# create pairwise scatter plot
pairs(classifications)
```



# output session information
```{r}
sessionInfo()
```



