---
title: "Data Smoothing (STAT5003 2016)"
author: "Pengyi Yang"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
#runtime: shiny
---

## Learning objective
The aim in this lecture is to understand the basic concepts in data smoothing etc.

## Data used in this lecture
(1) "simple data": created from a the function y = (x^3) * sin((x+3.4)/2)

(2) "tough data": created from two functions and concatenated together.

## Constant-span running mean smoothing
```{r}
# load data
# set the current directory to be current Rmd file
setwd("/Users/Sahana/Documents/STAT5003 Computational Statistical Methods/Smoothing")


source("functions.R")
easy <- read.table("easysmooth.dat", header=T)
x <- easy$X
y <- easy$Y

# The true relationship
s <- function(x){(x^3) * sin((x+3.4)/2)}

# plot data and create the true relationship line
x.plot <- seq(min(x),max(x),length.out=1000)
y.plot <- s(x.plot)
plot(x, y, xlab="Predictor", ylab="Response", main="Running mean smoothing")
lines(x.plot, y.plot, lty=2, lwd=2, col="red")

s3  <- movingAverage(3,y) # compute moving average with a neighbor of 3
s43 <- movingAverage(43,y) # compute moving average with a neighbor of 43

# plot moving average smoothing
lines(x, s3, col="blue3", lwd=1)
lines(x, s43, lty=5, col="green4", lwd=2)
legend("bottomright",c("True relation","k=3","k=43"),
      lty=c(2,1,3), col=c("red", "blue3", "green4"))

```

### Determining appropriate span using cross-validated residual sum of squares
```{r}
## CVRSS values of k = 3,5,7,...,51
k <- seq(3, 51, by = 2)

## Cross validation
cvrss.val <- rep(0,length(k))
for(i in 1:25){
      cvrss.val[i] <- CVRSS(k[i],y)
}

## plot cross validation results
plot(k,cvrss.val,type="b")
## output optimal k value
k[which.min(cvrss.val)]

# plot data and create the true relationship line
x.plot <- seq(min(x),max(x),length.out=1000)
y.plot <- s(x.plot)
plot(x, y, xlab="Predictor", ylab="Response", main="Running mean smoothing")
lines(x.plot, y.plot, lty=2, lwd=2, col="red")

# compute moving average with a CVRSS estimated optimal k value
s.hat <- movingAverage(k[which.min(cvrss.val)], y)

# plot moving average smoothing
lines(x, s.hat, col="purple", lwd=2)
legend("bottomright",c("True relation","CVRSS"),
      lty=c(2,1), col=c("red", "purple"))
```

## Running line smoothing
```{r}
# estimating s using running line smoother with different number of neighbours
s3 <- RLSmoother(3, y, x)
s23 <- RLSmoother(23, y, x)
s45 <- RLSmoother(45, y, x)

# true relationship
s <- function(x){(x^3) * sin((x+3.4)/2)}

# plot range
x.plot <- seq(min(x), max(x), length.out=1000)
y.plot <- s(x.plot)
plot(x, y, xlab="Predictor", ylab="Response", main="Running line smoothing")
lines(x.plot, y.plot, lty=2, lwd=2, col="red")
lines(x, s3, type="l", col="blue3", lwd=2)
lines(x, s23, type="l", col="purple", lwd=2)
lines(x, s45, type="l", col="green4", lwd=2)
legend("bottomright",c("True relation","k=3", "k=23", "k=45"),
      lty=c(2,1,1,1), col=c("red", "blue3", "purple" ,"green4"))
```

## Kernal smoothing
```{r}
# setup the plot of original data and true relationship
x.plot <- seq(min(x), max(x), length.out=1000)
y.plot <- s(x.plot)
plot(x, y, xlab="Predictor", ylab="Response", main="Running line smoothing")
lines(x.plot, y.plot, lty=2, lwd=2, col="red")

# apply kernal smoothing using "KernSmooth" package
library(KernSmooth)
fit <- locpoly(x, y, kernel="normal", bandwidth = 0.1)
lines(fit, col="blue3", lwd=2)
fit <- locpoly(x, y, kernel="normal", bandwidth = 0.5)
lines(fit, col="green4", lwd=2)

# plot smoothing line with estimated optimal bandwidth
lines(fit, col="purple", lwd=2)
legend("bottomright",c("True relation","bandwidth=0.1", "bandwidth=0.5"),
      lty=c(2,1,1), col=c("red", "blue3", "purple"))
```

## Spline smoothing
```{r}
## Use the smooth.spline function in R
cubicSpline1.fit <- smooth.spline(x=x, y=y, cv=FALSE, spar=0.5)
cubicSpline2.fit <- smooth.spline(x=x, y=y, cv=FALSE, spar=1)
cubicSpline3.fit <- smooth.spline(x=x, y=y, cv=TRUE)

## Out put smoothing line
s = function(x){(x^3) * sin((x+3.4)/2)}
x.plot = seq(min(x),max(x),length.out=1000)
y.plot = s(x.plot)
plot(x,y,xlab="Predictor",ylab="Response", main="Cubic spline smoothing")
lines(x.plot, y.plot, lty=2, lwd=2, col="red")
lines(cubicSpline1.fit, col="blue3", lwd=2)
lines(cubicSpline2.fit, col="purple", lwd=2)
lines(cubicSpline3.fit, col="green4", lwd=2)
legend("bottomright",c("True relation","spar=0.5", "spar=est", "spar=1"),
      lty=c(2,1,1,1), col=c("red", "blue3", "purple" ,"green4"))
```

## Loess smoothing
```{r}
# fitting loess with different spans
lo1.fit <- loess(y~x, span=0.1)
lo2.fit <- loess(y~x, span=0.8)

## Output smoothing results
s = function(x){(x^3) * sin((x+3.4)/2)}
x.plot = seq(min(x),max(x),length.out=1000)
y.plot = s(x.plot)
plot(x, y, xlab="Predictor",ylab="Response", main="Loess smoothing")
lines(x.plot, y.plot, lty=2, col="red", lwd=2)
lines(x, predict(lo1.fit), col="blue2", lwd=2, lty=3)
lines(x, predict(lo2.fit), col="green4", lwd=2)
legend("bottomright",c("True relation","span=0.1", "span=0.8"),
      lty=c(2,3,1), col=c("red", "blue2","green4"))
```


## Super smoothing
```{r, warning=FALSE, message=FALSE, fig.width=12, fig.height=6}
setwd("/Users/Sahana/Documents/STAT5003 Computational Statistical Methods/Smoothing")
easy <- read.table("toughsmooth.dat", header=T)
x <- easy$X
y <- easy$Y
# create different smoothing parameters for runing line smoothing
n <- length(y)
h <- c(0.05*n, 0.2*n, 0.5*n)

ssmooth.val  = supsmu(x,y) # use R built-in supersmoother
rlsmooth.vals = mapply(RLSmoother, h+1, MoreArgs = list(y,x)) # apply running line smoothing with different smoothing parameters

## Setup plots
par(mfrow=c(1,2))
plot(x,y,xlab="Predictor",ylab="Response",main="Running-line Smooths")
lines(x,rlsmooth.vals[,1],type="l", col="blue2")
lines(x,rlsmooth.vals[,2],type="l",lty=2, col="orange3", lwd=2)
lines(x,rlsmooth.vals[,3],type="l",lty=3, col="green4", lwd=2)
legend("bottomright",c("k=0.05n","k=0.2n","k=0.5n"), col=c("blue2", "orange3", "green4"), lty=1:3)

plot(x,y,xlab="Predictor",ylab="Response",main="Supersmooth")
lines(ssmooth.val, col="purple", lwd=2)
```

# Output session information
```{r, echo=FALSE}
sessionInfo()
```





