## Installed package ClueR
#Phosphorylation is an essential protein post-translational modification characterized by the precise and #reversible addition of a phosphate group, by proteins called 'kinases', to their targets, called #'substrates' (Hunter, 1995). 

```{r}
library(ClueR)
```

## Installed the package hES and Dimension displayed below

```{r}
data(hES)
dim(hES)
head(hES)
```
## Transposed the hES data for creating the hierarchical clustering with respect to times
```{r}
trans.hES=t(hES)
dim(trans.hES)
```

## R code for hierarchical clustering of hES data and plot the clustering Dendrogram
```{r}
dist.hES=dist(trans.hES,method = "euclidean")
hc.hES=hclust(dist.hES,method="ave")
plot(hc.hES)
```

##Load the library e1071 and standardize the data to be unit free
```{r}
library(e1071)
standardize=function(mat)
{
means=apply(mat,1,mean)
stds=apply(mat,1,sd)
tmp=sweep(mat,1,means,FUN="-")
mat.stand=sweep(tmp,1,stds,FUN="/")
return(mat.stand)
}
hES.scaled=standardize(hES)
head(hES.scaled)
```

##performing clustering using the scaled data
```{r}
fc=cmeans(hES.scaled,centers = 9)
fuzzPlot(hES.scaled, fc, mfrow = c(3, 3))
```

##Clustering results validation using DunnIndex for Kmeans
```{r}
library(clValid)
dunn.ind <- clValid(hES.scaled[,-3], nClust=4:14, validation="internal", clMethods=c("kmeans"),maxitems=4000)
summary(dunn.ind)
optimalScores(dunn.ind)
par(mfrow=c(2,2))
plot(dunn.ind)
```

##Stability based metrics for validation of clusters
```{r}
library(clValid)
APN.ind <- clValid(hES.scaled[,-3], nClust=4:14, validation="stability", clMethods=c("kmeans"),maxitems = 4000)
summary(APN.ind)
optimalScores(APN.ind)
par(mfrow=c(2,2))
plot(APN.ind)
```
##Is k=9 the best choice of k?
As per the Dunn Index score, the optimum choice of k is 4. and as per the stability based validation index the optimum choice of k is also 4. This by both measures the best K for clustering using kmeans is 4.
